Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Cuts/edits to get paper within 16 page limit.

\section{Recursion Schemes}\label{sec:category-theory}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

Furthermore, folds and unfolds can be shown to capture a wide range of recursion
schemes such as primitive recursion, mutual recursion, dynamic programming
algorithms, polymorphic recursion, and recursion with accumulators. However, for
divide-and-conquer algorithms, folds need to be generalised to hylomorphisms,
which provide the ultimate basic building block for any other recursion scheme.
To do this, we look at recursion schemes from the point of view of category
theory.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
(Section~\ref{sec:coq-coalg}) and then we put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness~\ref{sec:coq-rec-coalgebras}. The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity\footnote{The QSort definition and correctness proof in the
equations
package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

>>>>

\section{Recursion Schemes}\label{sec:category-theory}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes, which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

Furthermore, folds and unfolds can be shown to capture a wide range of recursion
schemes, such as primitive recursion, mutual recursion, dynamic programming
algorithms, polymorphic recursion, and recursion with accumulators. However, for
divide-and-conquer algorithms, folds need to be generalised to hylomorphisms,
which provide the ultimate basic building block for any other recursion scheme.
To do this, we look at recursion schemes from the point of view of category
theory.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
(Section~\ref{sec:coq-coalg}) and then we put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness~\ref{sec:coq-rec-coalgebras}. The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity.
%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
% \footnote{The QSort definition and correctness proof in the
% equations
% package is under 200LOC --
% \href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
% and other online proofs of correctness are of similar complexity
% \href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
